---
title: 'R Concepts Review'
subtitle: Rob Leonard (robleonard@tamu.edu) 
output:
  html_document: default
---
# 1) Income by Gender and Degree 
**1a Create Box Plots**
```{r}
inc_deg_data <- read.csv("inc_deg_data.csv")
attach(inc_deg_data)
boxplot(income~interaction(gender,degree), data=inc_deg_data,
        names=c("Female Arts","Male Arts","Female Science","Male Science"),
        xlab="Gender & Degree", ylab="Income")
```

**1-b Report the mean, median, standard deviation, and first and third quartiles of income.**
```{r}
summary(income)
cat("Standard Deviation of Income:", sd(income))
```
**1-c Repeat with income expressed in dollars (rather than 1,000’s of dollars).**
```{r}
summary(income*1000)
cat("Standard Deviation of Income:", sd(income*1000))
```
**1-d Repeat excluding the minimum and maximum values**
```{r}
inc_deg_data <- inc_deg_data[order(income),]
summary(inc_deg_data$income[2:99]*1000)
cat("Standard Deviation of Income excl. min/max:", sd(inc_deg_data$income[2:99]*1000))
detach(inc_deg_data)
```

# 2) Normal Distribution
**2a Simulate normal distribution values (100x5) and report the column means**
```{r}
set.seed(101)
Mat2 <- matrix(rnorm(500),nrow=100)
Mat2means <- apply(Mat2,2,mean)
Mat2means
```
**Redo using Matrix/Vector Arithmetic**
```{r}
vect1 <- rep(1,100)
Mat2ameans <- (1/100)*(vect1%*%Mat2) 
Mat2ameans
```
**2b Histogram of the row ranges**
```{r}
Mat2range <- apply(Mat2,1,max)-apply(Mat2,1,min)
hist(Mat2range, main="Histogram of Ranges", xlab="Range")
```

# 3) Gamma Distribution
**Histogram of Gamma Distribution Simulation with Overlayed Normal Distribution**
```{r}
par(mfrow=c(1,3))
HistGam <- function(n,nreps=1000,gshape=2,gscale=2) {
  gmean=gshape*gscale
  gvar=gshape*gscale*gscale
  temp<-rep(0,nreps)
  for (i in 1:nreps) temp[i]=mean(rgamma(n,shape=gshape,scale=gscale))
  hist(temp,probability=TRUE,xlab="Mean",main=paste("Histogram n=",n),
       breaks ="Sturges",ylim=c(0,1.4) )
  xnorm <- seq(min(temp),max(temp),length=nreps+1)
  lines(xnorm,dnorm(xnorm,mean=gmean,sd=sqrt(gvar/n)),col=4,lwd=2)
  }
HistGam(n=10) 
HistGam(n=30) 
HistGam(n=90)  
```

# Matrix Algebra in R
**4a Create matrices A and B and calculate column sums of A**
```{r}
set.seed(101)
A <- matrix(c(rnorm(12),runif(8,min=-2,max=2)),nrow=5,ncol=4,byrow=TRUE)
set.seed(102)
B <- matrix(rbeta(20,2,1),nrow=5,ncol=4,byrow=TRUE)
colSums(A)
```
**4b Calculate A+B and display certain elements**
```{r}
C<- A+B
C[4,2]
C[4,4]
```
**4c Matrix Multiplication and Transpose**
```{r}
D <- A%*%t(B) 
D[4,2]
D[4,4]
```
**4d Matrix inverse and determinants**
```{r}
E <- t(B)%*%A
solve(E)
det(E)
```




